---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# michigan2014

<!-- badges: start -->
<!-- badges: end -->

This package provides functions for calculating total species richness,
native species richness, mean C, native mean C, FQI, native FQI, and adjusted FQI 
based on Michigan's 2014 FQAI database. This package is meant to be a small test-drive 
of a larger, more comprehensive package for FQAI data bases across the USA.

## Installation

You can install the development version of michigan2014 from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ifoxfoot/michigan2014")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(michigan2014)

#the package comes with an example site assessment
plants_of_crooked_island <- crooked_island_site_assessment

head(plants_of_crooked_island)

#here is an example of all metrics wrapped into one
all_metrics(plants_of_crooked_island)
```

