---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fqacalc

## A Floristic Quality Assessment Calculator for R

<!-- badges: start -->
<!-- badges: end -->

This package provides functions for calculating Floristic Quality Assessment (FQA) metrics based on regional FQA databases that have been approved for use by the US Army Corps of Engineers. 

## Installation

You can install the development version of `fqacalc` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ifoxfoot/fqacalc")
```

```{r}
#attach packages required for this tutorial
library(fqacalc)
library(stringr)
library(dplyr)
```

## Package Data

`fqacalc` contains all regional FQA databases that have been either fully approved or approved with reservations for use by the US Army Corps of Engineers. By referencing these databases, the package knows what Coefficient of Conservatism (or C Value) to give each plant that the user inputs. Users can see a list of regional databases using the `db_names()` function, and specific FQA databases can be accessed using the `view_db()` function. 

```{r}
#view a list of all 51 available databases
head(db_names())

#store the Connecticut database as an object
colorado <- view_db("colorado_2020")

#view it
head(colorado)
```

`fqacalc` also comes with a real site assessment from Crooked Island, Michigan, downloaded from the [Universal FQA Calculator](https://universalfqa.org/). 

The data is called `crooked_island` and is used to demonstrate how the package works.

```{r}
#take a look at crooked island's plants
head(crooked_island)

#look at the documentation for the data (bottom right pane of R studio)
?crooked_island

#load dataset into local environment
crooked_island <- crooked_island
```

## Functions to see if your observations will be counted

`fqacalc` contains two functions that help the user understand how the data they input relates to the regional database: `accepted_entries()` and `unassigned_plants()`. `accepted_entries()` is a function that shows which observations in the input data frame were successfully matched to a regional database. To demonstrate how these data viewing functions work I'm going to add a mistake to the `crooked_island` data set.

```{r}
#introduce a typo
mistake_island <- crooked_island %>% 
  mutate(name = str_replace(name, "Abies balsamea", "Abies blahblah"))

#store accepted entries
accepted_entries <- accepted_entries(#this is the data
                                     mistake_island, 
                                     #'key' to join the data to regional list
                                     key = "name", 
                                     #this is the regional list
                                     db = "michigan_2014", 
                                     #include native AND non-native entries
                                     native = F) 

#view accepted entries
head(accepted_entries)
```

Now, when we use `accepted_entries()` to see which species were matched to the regional dataset, we can see that we got a message about the species 'ABIES BLAHBLAH' being discarded and we can also see that the accepted entries dataset we created only has 34 entries.

In some cases, plants from the user list can be matched to the regional database, but the plant is not associated with any C Value. This is usually because not enough is known about that plant. Plants that are matched but have no C Value will be excluded from FQI metric calculation but they can *optionally* be included in other metrics like species richness, relative cover, relative frequency, relative importance, and mean wetness, as well as any summarizing functions containing these metrics. This option is denoted with the `allow_no_c` argument. 

`unassigned_plants()` is a function that shows the user which plants have not been assigned a C Value.

```{r}
#To see unassigned_plants in action we're going to Montana! 

#first I'll create a df of plants to input
no_c_plants<- data.frame(name = c("ABRONIA FRAGRANS", 
                                  "ACER GLABRUM", 
                                  "ACER GRANDIDENTATUM", 
                                  "ACER PLATANOIDES"))

#then I'll create a df of unassigned plants
unassigned_plants(no_c_plants, key = "name", db = "montana_2017")

```

As you can see, two of these species have no C Values.

## Unweighted FQI Metrics

`fqacalc` also contains a variety of functions that calculate Total Species Richness, Native Species Richness, Mean C, Native Mean C, Total FQI, Native FQI, and Adjusted FQI. All of these functions eliminate duplicate species, species that cannot be found in the regional database, and species without a C Value.

#### Function Arguments

All of the metric functions have the same arguments. (and two don't have the native argument)

* **x**: A data frame containing a list of plant species. This data frame *must* have one of the following columns: `scientific_name` or `acronym`. 

* **key**: A character string representing the column that will be used to join the input `x` with the regional FQA database. If a value is not specified the default is `"name"`. `"name"` and `"acronym"` are the only acceptable values for key.

* **db**: A character string representing the regional FQA database to use. See `db_names()` for a list of potential values.

* **native**: native Boolean (TRUE or FALSE). If TRUE, calculate metrics using only native species.

Additionally, `species_richness()` and `all_metrics()` have an argument called `allow_no_c`. If `allow_no_c = TRUE` than species that are in the regional FQAI but don't have C Values will be included. If `allow_no_c` is FALSE, then these species will be omitted. This argument is also found in `mean_w()` and all the relative functions. 

#### Functions

```{r}
#total mean c
mean_c(crooked_island, key = "acronym", db = "michigan_2014", native = FALSE)

#native mean C
mean_c(crooked_island, key = "acronym", db = "michigan_2014", native = TRUE)

#total FQI
FQI(crooked_island, key = "acronym", db = "michigan_2014", native = FALSE)

#native FQI
FQI(crooked_island, key = "acronym", db = "michigan_2014", native = TRUE)

#adjusted FQI (always includes both natives and non-natives)
adjusted_FQI(crooked_island, key = "acronym", db = "michigan_2014")
```

And finally, `all_metrics()` prints all the metrics in a data frame format.

```{r}
all_metrics(crooked_island, key = "acronym", db = "michigan_2014")
```

Also, all the functions are documented with help pages.

```{r}
?all_metrics
```

## Cover-Weighted Functions

Cover-Weighted Functions calculate the same metrics but they are weighted by how abundant each species is. Therefore, the input data frame must also have a column named `cover` containing cover values. Cover values can be continuous (i.e. percent cover) or classed (i.e. using the Braun-Blanquet method).

Cover-Weighted Functions come in two flavors: those that allow duplicate entries and those that don't. Not allowing duplicate species observations works best for calculating plot-level metrics, where each species is counted once along with its total cover value. Allowing duplicates work best for transect-level metrics, where repeated plots along a transect may contain the same species. As a note, if `allow_duplicates = FALSE` in a cover-weighted function, any duplicates species will be counted once and their cover values will be added together.

Even if you allow duplicate species, if you indicate a plot id column using the `plot_id` argument, duplicate species in the same plots will be counted once and their cover values will be added together.

#### Function Arguments

Cover-Weighted Functions have a few additional arguments: 

* **cover_metric**: A character string representing the cover method used. Acceptable cover methods are: `"percent_cover"`, `"carolina_veg_survey"`, `"braun-blanquet"`, `"doubinmire"`, and `"usfs_ecodata"`. `"percent_cover"` is the default and is recommended because it is the most accurate.
* **allow_duplicates**: Boolean (TRUE or FALSE). If TRUE, allow duplicate entries of the same species. If FALSE, do not allow species duplication. Setting `allow_duplicates` to TRUE is best for calculating metrics for multiple plots/quadrants which potentially contain the same species. Setting `allow_duplicates` to FALSE is best for calculating metrics for a single plot/quadrant, where each species is entered once along with its total cover value.

#### Functions

```{r}
#first I'll make a hypothetical plot with cover values
plot <- data.frame(acronym  = c("ABEESC", "ABIBAL", "AMMBRE", "ANTELE"),
                   name = c("Abelmoschus esculentus", 
                            "Abies balsamea", "Ammophila breviligulata", 
                            "Anticlea elegans; zigadenus glaucus"),
                   cover = c(50, 4, 20, 30))

#now I'll make up a transect
transect <- data.frame(acronym  = c("ABEESC", "ABIBAL", "AMMBRE", 
                                    "ANTELE", "ABEESC", "ABIBAL", "AMMBRE"),
                      cover = c(50, 4, 20, 30, 40, 7, 60),
                      quad_id = c(1, 1, 1, 1, 2, 2, 2))

#cover mean c (no duplicates allowed!)
cover_mean_c(plot, key = "acronym", db = "michigan_2014", 
             native = FALSE, cover_metric = "percent_cover", 
             allow_duplicates = FALSE)

#transect mean c (duplicates allowed)
cover_mean_c(transect, key = "acronym", db = "michigan_2014", 
             native = FALSE, cover_metric = "percent_cover",
             allow_duplicates = TRUE)

#cover-weighted FQI (again, you can choose to allow duplicates or not)
cover_FQI(transect, key = "acronym", db = "michigan_2014", native = FALSE, 
          cover_metric = "percent_cover",
          allow_duplicates = TRUE)

#transect summary function (allows duplicates)
transect_summary(transect, key = "acronym", db = "michigan_2014")
```

There is also a plot summary function that summarizes plots along a transect. Data is input as a single data frame containing species per plot. This data frame must also have a column representing the plot the species was observed in. This column is then passed to an additional argument

* **plot_id**: A character string representing the name of the column in `x` that indicates which plot the species was observed in.

Because it is sometimes useful to calculate the total amount of bare ground or un-vegetated water in a plot, the user can also choose to include bare ground or water. To get this feature to work, the user must set another argument:

* **allow_non_veg**: Boolean (TRUE or FALSE). If TRUE, allow input to contain un-vegetated ground and un-vegetated water.

If `allow_non_veg` is true, the user can include species "UNVEGETATED GROUND" or "UNVEGETATED WATER" along with their plant species. They can also use acronyms "GROUND" or "WATER".

```{r}
#print transect to view structure of data
transect_unveg <- data.frame(acronym  = c("GROUND", "ABEESC", "ABIBAL", "AMMBRE", "ANTELE", "WATER", "GROUND", "ABEESC", "ABIBAL", "AMMBRE"),
                             cover = c(60, 50, 4, 20, 30, 20, 20, 40, 7, 60),
                             quad_id = c(1, 1, 1, 1, 1, 2, 2, 2, 2, 2))

#plot summary of a transect (dplicates are not allowed)
plot_summary(x = transect_unveg, key = "acronym", db = "michigan_2014", 
             cover_metric = "percent_cover", 
             plot_id = "quad_id")
```

## Relative Functions

Relative functions calculate relative frequency, coverage, and importance for each category. There is also a species summary function that produces a summary of each species' relative metrics in the data frame. Relative functions always allow duplicate species observations. They also allow ground and water to be included.

Relative functions have one additional argument which tells the functions what to calculate the relative value of: 

* **col** A character string equal to 'species', 'family', or 'physiog'. 

Relative functions do not have a native argument.

```{r}
#say I want to calculate the relative value of a tree

#relative frequency
relative_freq(transect, key = "acronym", db = "michigan_2014", 
              col = "physiog")

#relative cover
relative_cover(transect, key = "acronym", db = "michigan_2014", 
               col = "family", cover_metric = "percent_cover")

#relative importance
relative_importance(transect, key = "acronym", db = "michigan_2014", 
                    col = "species", cover_metric = "percent_cover")

#species summary
species_summary(transect, key = "acronym", db = "michigan_2014", 
                cover_metric = "percent_cover")

#physiognomy summary
physiog_summary(transect, key = "acronym", db = "michigan_2014", 
                cover_metric = "percent_cover")
```

## Wetness metric

`fqacalc` has one wetness metric, which calculates the mean wetness coefficient per site. The wetness coefficient is based off of the USFWS Wetland Indicator Status. Negative wetness coefficients indicate a stronger affinity for wetlands, while positive wetland coefficients indicate an affinity for uplands.

```{r}
#mean wetness
mean_w(crooked_island, key = "acronym", db = "michigan_2014")
```

## The End
